Hydra battles in Coq

### 2020-10-24 doc/hydras.tex
First draft of the chapter on Gamma0 (to be reviewed).

\begin{remark}
The alternate proof of well-foundedness using \'Evelyne Contejean's work on recursive path ordering~\cite{DershowitzRPO, a3pat} is available in the library \href{../src/html/hydras.Epsilon0.Epsilon0rpo.html}{Epsilon0rpo}.
 \end{remark}

@@

\begin{Coqsrc}
Instance Epsilon0 : ON Lt compare.  
(* ... *)
\end{Coqsrc}

@@

@@

\chapter{The ordinal \texorpdfstring{$\Gamma_0$}{Gamma0}}

@@

\subsubsection*{Proof of $\psi(1,0)\times 13+ \omega < \psi(0,\psi(2,1))$}

@@

\begin{Coqsrc}
Lemma  nf_finite i : nf (fin (S i)).
Proof.
   constructor 2; auto with T2.
Qed.

@@

Example Ex8: nf (gcons 2 1 42 epsilon0).
Proof.
  constructor 3; auto with T2.
  - apply Ex4.
  - apply nf_finite.
  - apply nf_finite.
Qed.

\end{Coqsrc}
\chapter{Appendices}

>>>>

\begin{remark}
\label{remark:a3pat}
The alternate proof of well-foundedness using \'Evelyne Contejean's work on recursive path ordering~\cite{DershowitzRPO, a3pat} is available in the library \href{../src/html/hydras.Epsilon0.Epsilon0rpo.html}{Epsilon0rpo}.
 \end{remark}

@@

\label{instance-epsilon0}
\begin{Coqsrc}
Instance Epsilon0 : ON Lt compare.  
(* ... *)
\end{Coqsrc}

@@

\label{sect:phi-schutte}

@@

\chapter{The ordinal \texorpdfstring{$\Gamma_0$}{Gamma0} first draft)}

@@

\subsubsection*{Proof of $\psi(1,0)\times 13+ \omega < \psi(0,\psi(2,1))$}
\label{sect:ex6-first-proof}

@@

\begin{Coqsrc}
Lemma  nf_fin i : nf (fin i).
Proof.
  destruct i.
  - auto with T2.
  - constructor 2; auto with T2.
Qed.

Lemma nf_omega : nf omega.
Proof.  compute; auto with T2. Qed.

@@

Lemma nf_epsilon : forall alpha, nf alpha -> nf (epsilon alpha).
Proof. compute; auto with T2. Qed.

Example Ex8: nf (gcons 2 1 42 epsilon0).
Proof.
  constructor 3; auto with T2.
  - apply Ex4.
  - apply nf_fin.
  - apply nf_fin.
Qed.
\end{Coqsrc}

\subsection{Length of a term}

The notion of \emph{term length} is introduced by Schütte as a helper for proving (at least) the \emph{trichotomy} property and transitivity of the strict order \texttt{lt} on \texttt{T2}. These properties are proved by induction on length.

\begin{Coqsrc}
Fixpoint nbterms (t:T2) : nat :=
  match t with zero => 0
             | gcons a b n v => (S n) + nbterms v
  end.

Fixpoint t2_length (t:T2) : nat :=
  match t  with 
    zero => 0
  | gcons a b n v => 
       nbterms (gcons a b n v) + 
      2 * (Max.max (t2_length a)
                              (Max.max (t2_length b) 
                                                (t2_length_aux v)))
  end
with t2_length_aux (t:T2) : nat :=
 match t with zero => 0
            | gcons a b n v =>
               Max.max (t2_length a) 
                                (Max.max (t2_length b) (t2_length_aux v))
 end.
\end{Coqsrc}

\begin{Coqsrc}
Compute t2_length (gcons 2 1 42 epsilon0).
\end{Coqsrc}

\begin{Coqanswer}
 = 48 : nat
\end{Coqanswer}

\subsection{Trichotomy}

\emph{Trichotomy} is another name for the well-known property of decidable total ordering (like Standard Library's \texttt{Compare\_dec.lt\_eq\_lt\_dec}).

We first prove by induction on $l$ the following lemma:

\noindent\emph{From \href{../src/html/hydras.Gamma0.Gamma0\#tricho_aux}%
{\texttt{Gamma0.Gamma0}}}

\begin{Coqsrc}
Lemma tricho_aux (l: nat) : forall t t' :T2,
      t2_length t + t2_length t' < l  ->
      {t t2< t'} + {t = t'} + {t' t2<  t}.
\end{Coqsrc}

Then we get our version of \texttt{lt\_eq\_lt\_dec}, and derive a comparison function;

\begin{Coqsrc}
Definition lt_eq_lt_dec (t t': T2) : {t t2< t'}+{t = t'}+{t' t2<  t}.
Proof.
  eapply tricho_aux.
  eapply lt_n_Sn.
Defined.

Definition compare (t1 t2 : T2) : comparison := 
  match lt_eq_lt_dec t1 t2 with
  | inleft (left _) => Lt
  | inleft (right _) => Eq
  | inright _ => Gt
  end.
\end{Coqsrc}

With the help of \texttt{compare}, we get a boolean version of \texttt{nf}
(being in Veblen normal form).

\begin{Coqsrc}
Fixpoint nfb (alpha : T2) : bool :=
  match alpha with
    zero => true
  | gcons a b n zero => andb (nfb a) (nfb b)
  | gcons a b n ((gcons a' b' n' c') as c) =>
    match compare [a', b'] [a, b] with
           Lt => andb (nfb a) (andb (nfb b) (nfb c))
           | _ => false
           end
end.
\end{Coqsrc}

\begin{Coqsrc}
Compute compare (gcons 2 1 42 epsilon0) [2,2].
\end{Coqsrc}

\begin{Coqanswer}
   = Lt
     : comparison
\end{Coqanswer}

\begin{Coqsrc}
Compute nfb  (gcons 2 1 42 epsilon0).
\end{Coqsrc}

\begin{Coqsrc}
Compute nfb (gcons 2 1 42 (gcons 2 2 4 epsilon0)).
\end{Coqsrc}

\begin{remark}
The connexion between the predicate \texttt{nf} and the relation \texttt{lt} on one part, and the functions \texttt{nfb} and \texttt{compare} on the other, is expressed by the following lemmas:

\begin{Coqsrc}
Lemma nfb_equiv gamma : nfb gamma = true <-> nf gamma.

The function \texttt{compare} helps to make easier proofs of inequalities of
closed terms of type \texttt{T2}.

First, we prove a lemma:

\begin{Coqsrc}
Lemma compare_Lt : forall alpha beta, compare alpha beta = Lt -> 
                                         alpha t2< beta.
Proof.
  intros alpha beta; destruct (compare_correct alpha beta);
    trivial; discriminate. 
Qed.
\end{Coqsrc}

Then, we give another version of the proof of Sect.~\vref{sect:ex6-first-proof}.

\begin{Coqsrc}
Example Ex6 : gcons 1 0 12 omega t2< [0,[2,1]].
Proof. now apply compare_Lt. Qed.
\end{Coqsrc}

\section{Main functions on \texttt{T2}}

\subsection{Successor}
The successor function is defined by structural recursion.

\noindent\emph{From \href{../src/html/hydras.Gamma0.T2.html\#succ}%
{\texttt{Gamma0.T2}}}
\begin{Coqsrc}
Fixpoint succ (a:T2) : T2 :=
 match a with zero => one
             | gcons zero zero n c => fin (S (S n))
             | gcons a b n c => gcons a b n (succ c)
 end.
\end{Coqsrc}

\subsection{Addition}

Like for Cantor normal forms (see Sect.\~ref{sect:infix-plus-T1})  the definition of addition in \texttt{T2}  requires comparison between ordinal terms.

\begin{Coqsrc}
Fixpoint plus (t1 t2 : T2) {struct t1}:T2 :=
  match t1,t2 with
  |  zero, y  => y
  |  x, zero => x
  |  gcons a b n c, gcons a' b' n' c' =>
     (match compare (gcons a b 0 zero)
                    (gcons a' b' 0 zero)
      with | Lt => gcons a' b' n' c'
      | Gt =>
        (gcons a b n
              (c +
               (gcons a' b' n' c')))
      | Eq  => (gcons a b (S(n+n')) c')
      end)
  end
where "alpha + beta" := (plus alpha beta): T2_scope.
\end{Coqsrc}

\begin{Coqsrc}
Example Ex7 : 3 + epsilon0 = epsilon0.
Proof. trivial. Qed.
\end{Coqsrc}

\subsection{The Veblen function \texorpdfstring{$\varphi$}{\texttt{phi}}}

The enumeration function of critical ordinals, presented in Sect.\~vref{sect:phi-schutte}, is recursively defined in type \texttt{T2}.

\begin{Coqsrc}
Definition  phi (alpha beta : T2) : T2 :=
  match beta with zero => [alpha, beta] 
             | [b1, b2] => 
               (match compare alpha b1
                with Datatypes.Lt => [b1, b2 ]
                | _ => [alpha,[b1, b2]]
                end)
             | gcons b1 b2 0 (gcons zero zero  n zero) => 
               (match compare alpha b1
                with  Datatypes.Lt => 
                      [alpha, (gcons b1 b2 0 (fin n))]
                | _ =>  [alpha, (gcons b1 b2 0 (fin (S n)))]
                end)
             | any_beta => [alpha, any_beta]
  end.
\end{Coqsrc}

Despite its complexity, the function \texttt{phi} is well adapted to proofs by simplification or computation.
\begin{Coqsrc}
Example Ex8:  phi 1 (succ epsilon0) = [1, [1,0] + 1].
Proof. reflexivity. Qed.
\end{Coqsrc}

\begin{Coqsrc}
(**  All epsilons are fixpoints of phi 0 *)

Theorem epsilon_fxp : forall beta, phi zero (epsilon beta) =
                                   epsilon beta.
Proof. reflexivity. Qed.

Theorem epsilon0_fxp : epsilon0 = phi zero epsilon0.
Proof. apply epsilon_fxp. Qed.
\end{Coqsrc}

The relation between the constructor $\psi$ and the function $\phi$ is
studied in~\cite{schutte}, and partially implemented in this development.
\emph{Please contribute!}

For instance, the following theorem states that, if $\gamma$ is the sum of a limit ordinal $\beta$ and a finite ordinal $n$, and $\beta$ is a fixpoint of
$\varphi(\alpha)$, then $\psi(\alpha,\gamma)=\phi(\alpha, \gamma+1)$.

\begin{Coqanswer}
phi_psi :
forall (alpha : T2) [beta gamma : T2] [n : nat],
nf gamma ->
limit_plus_fin beta n gamma ->
phi alpha beta = beta -> [alpha, gamma] = phi alpha (succ gamma)
\end{Coqanswer}

\begin{Coqsrc}
Example Ex9 : [zero, epsilon 2 + 4] = phi 0 (epsilon 2 + 5).
Proof. trivial. Qed.
\end{Coqsrc}

On the other hand, $\varphi$ can be expressed in terms of $\psi$.

\begin{Coqanswer}
phi_of_psi:
  forall a b1 b2 : T2,
  phi a [b1, b2] = (if lt_ge_dec a b1 then [b1, b2] else [a, [b1, b2]])
\end{Coqanswer}

\begin{Coqsrc}
Example Ex10 : phi omega [epsilon0, 5] = [epsilon0, 5].
Proof. reflexivity. Qed.
\end{Coqsrc}

\section{An ordinal notation for \texorpdfstring{$\Gamma_0$}{\texttt{Gamma0}}}

In order to consider type \texttt{T2} as an ordinal notation, we have to build an instance of class \texttt{ON} (See Definition page~\pageref{types:ON}).

First, we define a type that contains only terms in Veblen normal form, and redefine \texttt{lt} and \texttt{compare} by delegation.

\begin{Coqsrc}
Class G0 := mkg0 {vnf : T2; vnf_ok : nfb vnf}.

Definition lt (alpha beta : G0) := T2.lt (@vnf alpha) (@vnf beta).

Definition compare alpha beta := Gamma0.compare (@vnf alpha) (@vnf beta).
\end{Coqsrc}

Then we, prove that \texttt{lt} is a well-founded strict order and that the
function \texttt{compare} is correct.

\begin{Coqsrc}
Instance lt_sto : StrictOrder lt.

Instance Gamma0: ON lt  compare.
Proof.
  split.
  - apply lt_sto.
  - apply lt_wf. 
  - apply compare_correct.
Qed.
\end{Coqsrc}

\begin{remark}
The proof of \texttt{lt\_wf} has been written by \'Evelyne Contejean, using her library on the recursive path ordering (see also remark~\vref{remark:a3pat}).
\end{remark}

\index{Projects}
\begin{project}
Prove that \texttt{Epsilon0} (page~\pageref{instance-epsilon0})
is a sub-notation system of \texttt{Gamma0}.

Prove that the implemantations of \texttt{succ}, \texttt{+}, $\varphi_0$, etc.
are compatible in both notation systems.

Note that a function \texttt{T1\_inj} from \texttt{T1} to \texttt{T2} has already been defined. It may help to complete the task.
\end{project}

\chapter{Appendices}
